Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-09-13 Code/Manuscript.Rmd
updating

### Model occurrence

```{r select plots to control for sampling effort, eval=T}
#df.sub <- df[df$nplots>5, ]
df.sub <- df
## remove plots with other hosts
df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
#nrow(datBB.c)
df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]
```

```{r select plots with large & abundant, eval=T}
PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.10)
PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.10)
ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.10)

PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.10)
PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.10)
ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.10)
n
# PICO, PIEN & ABLA
df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.PIEN.ABLA$focalhosts2 <- "PICO, PIEN & ABLA"
df.sub.PICO.PIEN.ABLA$nhosts2 <- 3

# PICO & PIEN
df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]

df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]

df.sub.PICO.PIEN$focalhosts2 <- "PICO & PIEN"
df.sub.PICO.PIEN < - df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
df.sub.PICO.PIEN$nhosts2 <- 2

# PICO & ABLA
df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.ABLA$focalhosts2 <- "PICO & ABLA"
df.sub.PICO.ABLA$nhosts2 <- 2

df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

# PIEN & ABLA
df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PIEN.ABLA$focalhosts2 <- "PIEN & ABLA"
df.sub.PIEN.ABLA$nhosts2 <- 2

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

###
df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)

## Affected hosts
affectedhosts2 <- df.focal2 %>% group_by(focalhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhosts2[affectedhosts2$affected=="affected",]
affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts2[affectednhosts2$affected=="affected",]
### MPB

MPBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))

### PIEN

### ABLA

```{r, eval=T}
datBB.c <- df.sub[df.sub$focalhost=="Any",]
splitit <- initial_split(df.sub, prop=0.5)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, v=10)
```

```{r fit MPB RF model, eval=T}
response <- "agentMPB"
#vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO", "pressureMPB", "hostBA.MPB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

@@

# tune model
tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))

@@

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-MPB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-MPB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-MPB-occurrence.csv"), row.names=F)

@@

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-MPB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var", which.class="MPB")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-MPB-occurrence.csv"), row.names=F)
```

```{r fit SB RF model, eval=T}
response <- "agentSB"
#vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN", "pressureSB", "hostBA.SB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

@@

# Use backward selection to identify important variables
dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
registerDoSEQ()
vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-SB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-SB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-SB-occurrence.csv"), row.names=F)

@@

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-SB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var", which.class="SB")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-SB-occurrence.csv"), row.names=F)
```

```{r fit WBBB RF model, eval=T}
response <- "agentWBBB"
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA", "pressureWBBB", "hostBA.WBBB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

@@

# Use backward selection to identify important variables
dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
registerDoSEQ()
vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-WBBB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-WBBB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-WBBB-occurrence.csv"), row.names=F)

@@

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-WBBB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="WBBB")
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-WBBB-occurrence.csv"), row.names=F)
```

```{r, eval=F}
mpbo <- read.csv(here("Results/RF-pred-MPB-occurrence.csv"))
sbo <- read.csv(here("Results/RF-pred-SB-occurrence.csv"))
wbbbo <- read.csv(here("Results/RF-pred-WBBB-occurrence.csv"))

bbo <- datz.test
bbo$mpbpred <- mpbo$.pred_MPB
bbo$sbpred <- sbo$.pred_SB 
bbo$wbbbpred <- wbbbo$.pred_WBBB
bbo$mpba<- mpbo$agentMPB
bbo$sba<- sbo$agentSB
bbo$wbbba<- wbbbo$agentWBBB

bbo.sub <- bbo[bbo$mpbpred>0.25 &bbo$wbbbpred>0.25 & bbo$sbpred>0.25,   ]
nrow(bbo.sub)
bbo$actualSUM <- ifelse(bbo$mpba=="absent",0, 1) + ifelse(bbo$sba=="absent",0, 1) + ifelse(bbo$wbbba=="absent",0, 1)

bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)

bbo$pres <- factor(ifelse(bbo$actualSUM >0, 1, 0), "affected", "unaffected")

ggplot(bbo, aes(x=predSUM, y=actualSUM))+geom_boxplot()

bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
bbo

bbo$rMPB <- ifelse(bbo$agentMPB=="MPB", 1,0) - bbo$mpbpred 
bbo$rSB <- ifelse(bbo$agentSB=="SB", 1,0) - bbo$sbpred 
bbo$rWBBB <- ifelse(bbo$agentWBBB=="WBBB", 1,0) - bbo$wbbbpred
bbo$rSUM <- bbo$rMPB + bbo$rSB + bbo$rWBBB

x <- read.csv(here("Results/RF-partialdependence-SB-occurrence.csv"))
```

#### Severity
##### MPB
##### SB
##### WBBB

>>>>

```{r}
Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
df$hostsFPshort<- df$hostsFPlong
df$hostsFPshort <- factor(df$hostsFPlong, levels=levels(df$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
df$agentsFplongrev <- df$agentsFplong
df$agentsFplongrev <- factor(df$agentsFplongrev, levels=rev(levels( df$agentsFplong)))
jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
ggplot(df[df$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
whatever <- dev.off()
```

### Given suitable stand conditions, is co-occurrence common?

Construct random forest models using synthetic minority oversampling technique (SMOTE)

#### MPB

```{r fit MPB RF model, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")

@@

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))

@@

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
#registerDoSEQ()
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars
BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)

# generate a confusion matrix for test set
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)

@@

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
```

#### SB
```{r fit SB RF model, eval=F}
response <- "agentSB"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")

@@

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
#registerDoSEQ()
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars
BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)

# generate a confusion matrix for test set
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)

@@

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
```

#### WBBB
```{r fit WBBB RF model, eval=F}
response <- "agentWBBB"
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")

@@

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
#registerDoSEQ()
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars
BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)

# generate a confusion matrix for test set
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)

@@

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
```

```{r}
MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
MPBpp$focalagent <- "MPB"
SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
SBpp$focalagent <- "SB"
WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
WBBBpp$focalagent <- "WBBB"

ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"

x <- df[!df$focalagent=="Any",]
x <- x[,c("QMD.host", "propTBA", "BA.host", "focalagent", "affected01")]
x$affected01 <- ifelse(x$affected01=="affected", 1, 0)

xm <- melt(x, id.vars=c("focalagent", "affected01"))
colnames(xm) <- c("focalagent", "yhat", "var", "value")
xm$var <- factor(xm$var, levels=c("QMD.host", "propTBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
xm$focalagent <- as.character(xm$focalagent)
p <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+geom_point(data=xm[xm$var=="BA",], size=0, )+ylim(0.1,0.6)
ggplot(xm[xm$var=="BA",],aes(x=value, y=yhat, col=focalagent)

ggplot(xm, aes(x=focalagent, y=value))
```

```{r}

mpb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="MPB","affected"], bbo.dat[bbo.dat$focalagent=="MPB","hostspred"]))
sb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="SB","affected"], bbo.dat[bbo.dat$focalagent=="SB","hostspred"]))
wbbb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="WBBB","affected"], bbo.dat[bbo.dat$focalagent=="WBBB","hostspred"]))
any<- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="Any","affected"], bbo.dat[bbo.dat$focalagent=="Any","hostspred"]))

results <- NULL
for(j in levels(bbo.dat$hostspred)[5:8]){
  anyj <- any[any$Var2==j,]
  mpbj <- mpb[mpb$Var2==j,]
  sbj <- sb[sb$Var2==j,]
  wbbbj <- wbbb[wbbb$Var2==j,]

  x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))

  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$hostspred <-j
  res$barkbeetle <- "MPB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)

    x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))

  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$hostspred <-j
  res$barkbeetle <- "SB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)

   x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")

  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$hostspred <-j
  res$barkbeetle <- "WBBB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)
}

results$p <- round(results$p,3)

x <- bbo.dat[is.na(bbo.dat$hostspred)==FALSE,]
x <- x[x$hostspred %in% levels(x$hostspred)[5:8],]
x[x$hostspred == "PICO\n& PIEN" & x$focalagent=="WBBB","affected"]<- 0
x[x$hostspred == "PICO\n& ABLA" & x$focalagent=="SB","affected"]<- 0
x[x$hostspred == "PIEN\n& ABLA" & x$focalagent=="MPB","affected"]<- 0

xx <- x %>% group_by(hostspred,focalagent) %>% count(affected)
xxx <- x %>% group_by(hostspred) %>% count()

xxxx <- merge(xx, xxx, by="hostspred")
xxxx$prop <- xxxx$n.x/xxxx$n.y
xxxx <- xxxx[xxxx$affected=="affected",]
xxxx <- xxxx[,c(1,2,6)]
colnames(xxxx) <- c("hostspred", "barkbeetle", "prop")
